Variants of G\"{o}del's Ontological Proof \\ in a Natural Deduction Calculus

### 2016-04-06 Papers/InProgress/NaturalDeduction/GodProof-ND.tex
Add numbers and references to subproofs of Bjordal's variant
Add missing definition folding in Bjordal's proof
Make MCP a new command, in order to avoid spurious spaces in math mode
Improve section about Anderson's proof without superfluous axioms

We know from \citep{hajek} that Anderson's emendation also makes Axioms 4 and 5 derivable from the other axioms and definitions. Discussions about the redundancy of these two axioms started with an earlier (and wrong claim) claim by Magari \citep{magari} w.r.t. Scott's variant, and the redundancy was the topic of a long controversy \citep{controversy} between H\'ajek and Anderson. The proof presented above still uses Axioms 4 and 5, though, since the intention here was to show that Anderson's emendation requires only a minimal modification of the proofs presented in the previous sections.

\subsection{A shorter proof without redundant axioms}

@@

There is an alternative proof of theorem \ref{A:T3} using only Anderson's axioms \ref{A:A1}, \ref{A:A2} and \ref{A:A3} as well as definition \ref{A:D1}. %Hájek1996 could be helpful.
The proof requires only a modification of the proof of lemma \ref{A:L1}. The rest of the dependencies are as in the previous proof.

Theorem \ref{A:T1} is derived from axiom \ref{A:A1} and axiom \ref{A:A2}. Corollary \ref{A:C1} is derived from axiom \ref{A:A3} and theorem \ref{A:T1}. The proof of lemma \ref{A:L1} can be modified, so as to be derived from axiom \ref{A:A3} using definition \ref{A:D1}. Finally, theorem \ref{A:T3} can be derived from lemma \ref{A:L1} and corollary \ref{A:C1}.

@@

\begin{scriptsize}
\begin{prooftree}
\AXC{$\subproof_6$} \dashedLine
\UIC{$G(\gamma)$}  \dottedLine \RightLabel{D\ref{D1}}
\UIC{$\forall \varphi. [P(\lambda x.\varphi(x)) \biimp \nec \varphi(\gamma)]$}\RightLabel{$\forall_E$}
\UIC{$P(G) \biimp \nec G(\gamma)$}
\AXC{Axiom \ref{A:A3}} \RightLabel{}
\UIC{$P(G)$} \RightLabel{$\biimp_E$}
\BIC{$\nec G(\gamma)$}
\AXC{$ $} \RightLabel{5}
\UIC{$G(\gamma)$} \RightLabel{$\ex_I$}
\UIC{$\ex x.G(x)$}  \RightLabel{$\imp_I^5$}
\UIC{$G(\gamma)\imp \ex x.G(x)$} \RightLabel{$\nec_I$}
\UIC{$\nec [G(\gamma)\imp \ex x.G(x)]$}  \RightLabel{Axiom K}
\UIC{$\nec G(\gamma)\imp \nec \ex x.G(x)$} \RightLabel{$\imp_E$}
\BIC{$\nec \ex x.G(x)$}\RightLabel{$\imp_I^1$}
\UIC{$\ex z. G(z)\imp \nec \ex x.G(x)$}
\end{prooftree}
\end{scriptsize}

@@

@@

\begin{definition}[Changed]
\label{B:D2}
A \emph{maximal composite} of an individual's positive properties is a positive property possessed by the individual and necessarily implying every positive property possessed by the individual.
$$MCP(\varphi, x) \equiv (\varphi(x) \wedge P_B (\varphi)) \wedge \all\psi.((\psi(x) \wedge P_B (\psi)) \imp \nec\all y.(\varphi(y) \imp \psi(y)))
$$
\end{definition}

\begin{definition}[Changed]
\label{B:D3}
\emph{Necessary existence} of an individual is the necessary exemplification of all its maximal composites.
$$
NE_B (x) \equiv \all\varphi. (MCP(\varphi, x) \imp \nec \ex y.\varphi(y))
$$
\end{definition}

@@

\begin{lemma}
\label{B:L1}
If there is a God-like being, then there is a God-like being necessarily:
$$
\ex z. G_B(z) \imp \nec \ex x. G_B(x)
$$
\end{lemma}
\begin{proof}
\begin{small}
\begin{prooftree}
\AXC{$ $}
\UIC{$G_B(x) $} 
\AXC{$ $}\doubleLine
\UIC{$\nec \all x. [G_B(x) \imp G_B(x)] $}\RightLabel{ D\ref{B:D1}}\dottedLine
\UIC{$P_B(G_B) $}\RightLabel{ $\wedge_I$}
\BIC{$G_B(x) \wedge P_B(G_B) $}
\end{prooftree}
\end{small}

\begin{small}
\begin{prooftree}
\AXC{$ $}\dashedLine
\UIC{$G_B(x) \wedge P_B(G_B) $}
\AXC{$ $}\RightLabel{ $1$}
\UIC{$(\psi(x) \wedge P_B (\psi))$} \RightLabel{ $\wedge_E$}
\UIC{$P_B (\psi)$}\RightLabel{ D\ref{B:D1}}\dottedLine
\UIC{$\nec\all y.(G_B(y) \imp \psi(y))$} \RightLabel{ $\imp_I^1$}
\UIC{$(\psi(x) \wedge P_B (\psi))\imp \nec\all y.(G_B(y) \imp \psi(y))$}\RightLabel{ $\forall_I$}
\UIC{$\all\psi.((\psi(x) \wedge P_B (\psi)) \imp \nec\all y.(G_B(y) \imp \psi(y)))$} \RightLabel{ $\wedge_I$}
\BIC{$MCP(G_B, x) $}

@@

\begin{small}
\begin{prooftree}
\AXC{Axiom \ref{A:A5'}}\dashedLine
\UIC{$P_B (NE_B)$} \RightLabel{ D\ref{B:D1}}\dottedLine
\UIC{$\nec \all x. [G_B(x) \imp NE_B(x)]$}\RightLabel{ D\ref{B:D3}}\dottedLine
\UIC{$\nec \all x. [G_B(x) \imp \all\varphi. (MCP(\varphi, x) \imp \nec \ex y.\varphi(y))]$}  \RightLabel{ $\nec_E$}
\UIC{$\all x. [G_B(x) \imp \all\varphi. (MCP(\varphi, x) \imp \nec \ex y.\varphi(y))]$}  \RightLabel{ $\forall_E$}
\UIC{$ [G_B(x) \imp \all\varphi. (MCP(\varphi, x) \imp \nec \ex y.\varphi(y))]$}
\AXC{$ $} \RightLabel{ $2$}
\UIC{$G_B(x) $}  \RightLabel{ $\imp_E$}
\BIC{$\all\varphi. (MCP(\varphi, x) \imp \nec \ex y.\varphi(y))$} \RightLabel{ $\forall_E$}
\UIC{$(MCP(G_B, x) \imp \nec \ex y.G_B(y))$}
\AXC{$ $}\dashedLine
\UIC{$MCP(G_B, x) $} \RightLabel{ $\imp_E$}
\BIC{$\nec \ex y.G_B(y)$}\RightLabel{ $\imp_I^2$}
\UIC{$G_B(x) \imp \nec \ex y.G_B(y)$}
\end{prooftree}
\end{small}
\end{proof}

>>>>

H\'ajek \citep{hajek} claimed that Anderson's emendation makes Axioms 4 and 5 derivable from the other axioms and definitions. Discussions about the redundancy of these two axioms started with an earlier (and wrong claim) claim by Magari \citep{magari} w.r.t. Scott's variant, and the redundancy was the topic of a long controversy \citep{controversy} between H\'ajek and Anderson. As confirmed in Subsection \ref{sec:NonRedundantAnderson}, H\'ajek's claim is correct. Nevertheless, the proof presented above still uses Axioms 4 and 5, since this section's intention is to show that Anderson's emendation requires only a minimal modification of the proofs presented in the previous sections.

\subsection{A Shorter Proof without Superfluous Axioms}
\label{sec:NonRedundantAnderson}

@@

In order to construct a natural deduction proof of theorem \ref{A:T3} using only Anderson's axioms \ref{A:A1}, \ref{A:A2} and \ref{A:A3} as well as Anderson's Definition \ref{A:D1}, it suffices to modify 
the proof of Lemma \ref{A:L1}, which is the only part of the argument that depended on axioms \ref{A:A4} and \ref{A:A5} and definitions \ref{A:D2} and \ref{A:D3}. Such a modified proof is shown below.

@@

\begin{scriptsize}
\begin{prooftree}
\AXC{$\subproof_6$} \dashedLine
\UIC{$G(\gamma)$}  \dottedLine \RightLabel{D\ref{D1}}
\UIC{$\forall \varphi. [P(\lambda x.\varphi(x)) \biimp \nec \varphi(\gamma)]$}\RightLabel{$\forall_E$}
\UIC{$P(G) \biimp \nec G(\gamma)$}
\AXC{Axiom \ref{A:A3}} \RightLabel{}
\UIC{$P(G)$} \RightLabel{$\biimp_E$}
\BIC{$\nec G(\gamma)$}
\AXC{$ $} \RightLabel{5}
\UIC{$G(\gamma)$} \RightLabel{$\ex_I$}
\UIC{$\ex x.G(x)$}  \RightLabel{$\imp_I^5$}
\UIC{$G(\gamma)\imp \ex x.G(x)$} \RightLabel{$\nec_I$}
\UIC{$\nec [G(\gamma)\imp \ex x.G(x)]$} \doubleLine  \RightLabel{(using Axiom K)}
\UIC{$\nec G(\gamma)\imp \nec \ex x.G(x)$} \RightLabel{$\imp_E$}
\BIC{$\nec \ex x.G(x)$}\RightLabel{$\imp_I^1$}
\UIC{$\ex z. G(z)\imp \nec \ex x.G(x)$}
\end{prooftree}
\end{scriptsize}

@@

\noindent
Proofs of other lemmas, theorems and corollaries remain the same as in the previous section. Theorem \ref{A:T1} is derived from axioms \ref{A:A1} and \ref{A:A2}. Corollary \ref{A:C1} is derived from Axiom \ref{A:A3} and Theorem \ref{A:T1}. Finally, Theorem \ref{A:T3} can be derived from Corollary \ref{A:C1} and Lemma \ref{A:L1}, which, as shown above, depends only on Axiom \ref{A:A3} and Definition \ref{A:D1}. Therefore, confirming H\'ajek's claim, axioms \ref{A:A4} and \ref{A:A5} and definitions \ref{A:D2} and \ref{A:D3} are superfluous: they are not needed to prove Theorem \ref{A:T3}.

@@

\newcommand{\MCP}{\mathit{MCP}}

\begin{definition}[Changed]
\label{B:D2}
A \emph{maximal composite} of an individual's positive properties is a positive property possessed by the individual and necessarily implying every positive property possessed by the individual.
$$\MCP(\varphi, x) \equiv (\varphi(x) \wedge P_B (\varphi)) \wedge \all\psi.((\psi(x) \wedge P_B (\psi)) \imp \nec\all y.(\varphi(y) \imp \psi(y)))
$$
\end{definition}

\begin{definition}[Changed]
\label{B:D3}
\emph{Necessary existence} of an individual is the necessary exemplification of all its maximal composites.
$$
NE_B (x) \equiv \all\varphi. (\MCP(\varphi, x) \imp \nec \ex y.\varphi(y))
$$
\end{definition}

@@

\begin{lemma}
\label{B:L1}
If there is a God-like being, then there is a God-like being necessarily:
$$
\ex z. G_B(z) \imp \nec \ex x. G_B(x)
$$
\end{lemma}
\begin{proof}\hfill
\begin{small}
\begin{prooftree}
\AXC{$ $} \RightLabel{$2$}
\UIC{$G_B(x) $} 
\AXC{$ $}\doubleLine
\UIC{$\nec \all x. [G_B(x) \imp G_B(x)] $}\RightLabel{ D\ref{B:D1}}\dottedLine
\UIC{$P_B(G_B) $}\RightLabel{ $\wedge_I$}
\BIC{$[\subproof_{10}] \qquad G_B(x) \wedge P_B(G_B) $}
\end{prooftree}
\end{small}

\begin{small}
\begin{prooftree}
\AXC{$ \subproof_{10} $}\dashedLine
\UIC{$G_B(x) \wedge P_B(G_B) $}
\AXC{$ $}\RightLabel{ $1$}
\UIC{$(\psi(x) \wedge P_B (\psi))$} \RightLabel{ $\wedge_E$}
\UIC{$P_B (\psi)$}\RightLabel{ D\ref{B:D1}}\dottedLine
\UIC{$\nec\all y.(G_B(y) \imp \psi(y))$} \RightLabel{ $\imp_I^1$}
\UIC{$(\psi(x) \wedge P_B (\psi))\imp \nec\all y.(G_B(y) \imp \psi(y))$}\RightLabel{ $\forall_I$}
\UIC{$\all\psi.((\psi(x) \wedge P_B (\psi)) \imp \nec\all y.(G_B(y) \imp \psi(y)))$} \RightLabel{ $\wedge_I$}
\BIC{$G_B(x) \wedge P_B(G_B) \wedge \all\psi.((\psi(x) \wedge P_B (\psi)) \imp \nec\all y.(G_B(y) \imp \psi(y))) $} \RightLabel{ D\ref{B:D2} }\dottedLine
\UIC{$[\subproof_{11}] \qquad \MCP(G_B, x) $}

@@

\begin{small}
\begin{prooftree}
\AXC{Axiom \ref{A:A5'}}\dashedLine
\UIC{$P_B (NE_B)$} \RightLabel{ D\ref{B:D1}}\dottedLine
\UIC{$\nec \all x. [G_B(x) \imp NE_B(x)]$}\RightLabel{ D\ref{B:D3}}\dottedLine
\UIC{$\nec \all x. [G_B(x) \imp \all\varphi. (\MCP(\varphi, x) \imp \nec \ex y.\varphi(y))]$}  \RightLabel{ $\nec_E$}
\UIC{$\all x. [G_B(x) \imp \all\varphi. (\MCP(\varphi, x) \imp \nec \ex y.\varphi(y))]$}  \RightLabel{ $\forall_E$}
\UIC{$ [G_B(x) \imp \all\varphi. (\MCP(\varphi, x) \imp \nec \ex y.\varphi(y))]$}
\AXC{$ $} \RightLabel{ $2$}
\UIC{$G_B(x) $}  \RightLabel{ $\imp_E$}
\BIC{$\all\varphi. (\MCP(\varphi, x) \imp \nec \ex y.\varphi(y))$} \RightLabel{ $\forall_E$}
\UIC{$(\MCP(G_B, x) \imp \nec \ex y.G_B(y))$}
\AXC{$ \subproof_{11} $}\dashedLine
\UIC{$\MCP(G_B, x) $} \RightLabel{ $\imp_E$}
\BIC{$\nec \ex y.G_B(y)$}\RightLabel{ $\imp_I^2$}
\UIC{$G_B(x) \imp \nec \ex y.G_B(y)$}
\end{prooftree}
\end{small}
\end{proof}
